Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-11 paper/paper.tex
spell check.

When  there  are  multiple  occurences of  \bcar,  \bcdr\  and  \clazy,
$\hookrightarrow$  is  applied  from  right to  left.   The  reflexive
transitive   closure  of   $\hookrightarrow$   will   be  denoted   as
$\stackrel{*}{\hookrightarrow}$.   The  following proposition  relates
the original and the modified liveness rules.
\begin{proposition}
Assume that a liveness computation  based on the original set of rules
gives the liveness  of the variable $x$ at a  program point $\pi_i$ as
$\sigma$   (symbolically,   $\Lanv{i}{x}=   \sigma$).   Further,   let
$\Lanv{i}{x}=  \sigma'$ when the  modified rules  are used  instead of
\Lfunonly.  Then $\sigma' \stackrel{*}{\hookrightarrow} \sigma$.
\end{proposition}

@@

We circumvent the problem  of undecidability by over- approximating the
CFG by non-deterministic finite state automata (NFA). The NFA are then
simplified using  the $\hookrightarrow$ rules.  Finally the simplified
NFAs are converted to DFA.

@@

The strongly regular grammar is converted  into a set of NFAs, one for
each $\var{\Lanv{i}{x}}$.  The $\hookrightarrow$ simplification is now
done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
pairs  of consecutive  edges labeled  \bcar\acar\ or  \bcdr\acdr\ and
constructing the $\epsilon$-closure. Since the $\epsilon$-closure step
may  create  more  adjacent  \bcar\acar\ and  \bcdr\acdr\  edges,  the
simplification is continued till a fixed point is reached, after which
the edges labeled \bcar \ and  \bcdr are deleted.  The details of the
algorithm, its correctness and termination proofs are given by Karkare
et.  al.~\cite{karkare07liveness,asati14lgc}.  The resulting automaton
has  edges labeled  with  \acar,  \acdr\ and  \clazy\  only. In  this
automaton, for every edge labeled \clazy, we check if the source node
of the edge has a valid path to  a final final state.  If yes, we mark
the mark  the source node as  final. Finally, we remove  all the edges
labeled  \clazy\  and  convert  the automaton  into  a  deterministic
automaton.    This   effectively  implements   the   $\hookrightarrow$
simplification rules for  \bcar, \bcdr, and \clazy\  to obtain forward
paths. While  checking for liveness  during GC, a  forward access-path
only if it can reach a final state.
Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
from the grammar for \var{\Df{\length}{1}}.  The final DFA is shown in
Figure~\ref{fig:example-automata}(b).  This expectedly says that given
a  demand $\epsilon$  on  \length,  the liveness  of  its argument  is
$\acdr^{*}$  (the  spine  of   the  list  is  traversed).   Similarly,
Figure~\ref{fig:example-automata}(c)     shows     the     NFA     for
\var{\Lanv{9}{\pa}}.          The         DFA         shown         in
Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
paths.   This  reflects  the  lazy  nature  of  our  language.   Since
\length\ does not  evaluate the member elements of  the argument list,
the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
liveness based GC is triggered beyond $\pi_9$.

@@

The statistics related to compile  time liveness analysis are shown in
Table~\ref{tab:exp-results}(a). We observe that the analysis requires
a reasonable time except for  \verb@treejoin@ and  \verb@sudoku@. 
The bottleneck in our analysis is the NFA to DFA conversion
(worst-case exponential behaviour). To avoid this compile-time
overhead, we have added a caching mechanism in our simulator that can
reuse the DFAs from earlier analysis if the program has not changed..

Table~\ref{tab:exp-results}(b)   compares   the   garbage   collection
statistics  for RGC  and  LGC.  We  report the  number  of GC  events,
average  number of  cells reclaimed  per GC,  average number  of cells
touched  per GC,  total  number of  dragged cells  and  total time  to
perform  all collections.   It is  no  surprise that  number of  cells
reclaimed per  GC is  higher (and  consequently the  number of  GCs is
lower)  for LGC  based collection.   LGC  also reduces  the number  of
dragged cells significantly. However, the cost  for LGC is that per GC
execution  time is  more  for  LGC, which  results  in higher  overall
execution time  for garbage  collection, even  with reduced  number of
collections.   However, LGC  run time  is still  competitive (slowdown
within 5X of RGC for worst case), and is better for 3 benchmarks (best
case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
synthestic benchmark  that allocates complete binary  trees of various
sizes.  These binary trees are never used by the program. As a result,
the benchmark highly favours LGC.  We have included this benchmark for
completeness  only  and   the  numbers  for  it   are  not  considered
representative of practical programs.

@@

One area  of concern is the  huge gap between the  actual liveness and
the liveness  percieved by our  collector.  In  case of LGC  for eager
languages~\cite{asati14lgc} the gap was  very narrow and almost touched
the actual  liveness curve.  In  case of  lazy languages, this  gap is
largely   due  to   approximate  reachability   based  collection   of
closures. To implement liveness based  collection of closures, we need
to  record liveness  data  for  each closure  and  update  it at  each
evaluation  point.   These  space  and  run  time  overheads  make  it
infeasible  to  incorporate  liveness  based  GC  for  closures  in  a
practical garbage collector.

@@

  Although we  provide an  implementation for  a liveness  based garbage
  collection scheme for lazy language, we  do not provide a formal proof
  of its correctness.  A formal  proof of correctness would describe how
  closures  are  handled  during  garbage  collection  and  liveness  is
  propogated inside closures.

@@

%Add graphs for memort usage LGC vs RGC
%Add graphs/table showing the GC times and other statistics

>>>>

When  there  are  multiple  occurrences of  \bcar,  \bcdr\  and  \clazy,
$\hookrightarrow$  is  applied  from  right to  left.   The  reflexive
transitive   closure  of   $\hookrightarrow$   will   be  denoted   as
$\stackrel{*}{\hookrightarrow}$.   The  following proposition  relates
the original and the modified liveness rules.
\begin{proposition}
Assume that a liveness computation  based on the original set of rules
gives the liveness  of the variable $x$ at a  program point $\pi_i$ as
$\sigma$   (symbolically,   $\Lanv{i}{x}=   \sigma$).   Further,   let
$\Lanv{i}{x}=  \sigma'$ when the  modified rules  are used  instead of
\Lfunonly.  Then $\sigma' \stackrel{*}{\hookrightarrow} \sigma$.
\end{proposition}

@@

We circumvent the problem  of undecidability by over- approximating the
CFG by non-deterministic finite state automata (NFA). The NFA are then
simplified using  the $\hookrightarrow$ rules.  Finally the simplified
NFA are converted to DFA.

@@

The strongly regular grammar is converted  into a set of NFA, one for
each $\var{\Lanv{i}{x}}$.  The $\hookrightarrow$ simplification is now
done on the NFA by  repeatedly introducing $\epsilon$ edges to bypass
pairs  of consecutive  edges labeled  \bcar\acar\ or  \bcdr\acdr\ and
constructing the $\epsilon$-closure. Since the $\epsilon$-closure step
may  create  more  adjacent  \bcar\acar\ and  \bcdr\acdr\  edges,  the
simplification is continued till a fixed point is reached, after which
the edges labeled \bcar \ and  \bcdr are deleted.  The details of the
algorithm, its correctness and termination proofs are given by Karkare
et.  al.~\cite{karkare07liveness,asati14lgc}.  The resulting automaton
has  edges labeled  with  \acar,  \acdr\ and  \clazy\  only. In  this
automaton, for every edge labeled \clazy, we check if the source node
of the edge has a valid path to  a final final state.  If yes, we mark
the mark  the source node as  final. Finally, we remove  all the edges
labeled  \clazy\  and  convert  the automaton  into  a  deterministic
automaton.    This   effectively  implements   the   $\hookrightarrow$
simplification rules for  \bcar, \bcdr, and \clazy\  to obtain forward
paths. While  checking for liveness  during GC, a  forward access-path
only if it can reach a final state.
Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
from the grammar for \var{\Df{\length}{1}}.  The final DFA is shown in
Figure~\ref{fig:example-automata}(b).  This expectedly says that given
a  demand $\epsilon$  on  \length,  the liveness  of  its argument  is
$\acdr^{*}$  (the  spine  of   the  list  is  traversed).   Similarly,
Figure~\ref{fig:example-automata}(c)     shows     the     NFA     for
\var{\Lanv{9}{\pa}}.          The         DFA         shown         in
Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
paths.   This  reflects  the  lazy  nature  of  our  language.   Since
\length\ does not  evaluate the member elements of  the argument list,
the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
liveness based GC is triggered beyond $\pi_9$.

@@

The statistics related to compile  time liveness analysis are shown in
Table~\ref{tab:exp-results}(a). We observe that the analysis requires
a reasonable time except for  \verb@treejoin@ and  \verb@sudoku@. 
The bottleneck in our analysis is the NFA to DFA conversion
(worst-case exponential behaviour). To avoid this compile-time
overhead, we have added a caching mechanism in our simulator that can
reuse the DFA from earlier analysis if the program has not changed..

Table~\ref{tab:exp-results}(b)   compares   the   garbage   collection
statistics  for RGC  and  LGC.  We  report the  number  of GC  events,
average  number of  cells reclaimed  per GC,  average number  of cells
touched  per GC,  total  number of  dragged cells  and  total time  to
perform  all collections.   It is  no  surprise that  number of  cells
reclaimed per  GC is  higher (and  consequently the  number of  GCs is
lower)  for LGC  based collection.   LGC  also reduces  the number  of
dragged cells significantly. However, the cost  for LGC is that per GC
execution  time is  more  for  LGC, which  results  in higher  overall
execution time  for garbage  collection, even  with reduced  number of
collections.   However, LGC  run time  is still  competitive (slowdown
within 5X of RGC for worst case), and is better for 3 benchmarks (best
case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
synthetic benchmark  that allocates complete binary  trees of various
sizes.  These binary trees are never used by the program. As a result,
the benchmark highly favours LGC.  We have included this benchmark for
completeness  only  and   the  numbers  for  it   are  not  considered
representative of practical programs.

@@

One area  of concern is the  huge gap between the  actual liveness and
the liveness  perceived by our  collector.  In  case of LGC  for eager
languages~\cite{asati14lgc} the gap was  very narrow and almost touched
the actual  liveness curve.  In  case of  lazy languages, this  gap is
largely   due  to   approximate  reachability   based  collection   of
closures. To implement liveness based  collection of closures, we need
to  record liveness  data  for  each closure  and  update  it at  each
evaluation  point.   These  space  and  run  time  overheads  make  it
infeasible  to  incorporate  liveness  based  GC  for  closures  in  a
practical garbage collector.

@@

  Although we  provide an  implementation for  a liveness  based garbage
  collection scheme for lazy language, we  do not provide a formal proof
  of its correctness.  A formal  proof of correctness would describe how
  closures  are  handled  during  garbage  collection  and  liveness  is
  propagated inside closures.

@@

%Add graphs for memory usage LGC vs RGC
%Add graphs/table showing the GC times and other statistics
